---
title: "RNA seq Workflow with DeSEQ"
output: html_notebook
---
```{r}
getwd()
```
```{r}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
 if (!requireNamespace("BiocManager", quietly = TRUE))
     install.packages("BiocManager")
BiocManager::install("DESeq2", version = "3.14")

```
BiocManager::install("pheatmap")


# RNA-seq Analysis

workflow: raw counts -> normalization -> unsupervised clustering (PCA) -> DE analysis -> shrinking log2 fold change -> results exploration
- can add more downstream analysis as well

```{r}
BiocManager::install("pheatmap")
```

#dependencies
```{r}
library(DESeq2)
library(pheatmap)
```
# Preprocessing
## Reading and rounding the values of raw counts file (geneid x sample names)
```{r}
## Loading Input files and creating DESeq2 object
raw_counts <- read.delim("counts.txt", header = T, row.names = 1)
raw_counts <- ceiling(raw_counts) #round them up to integers
head(raw_counts)
```
## Loading metadata
```{r}
meta_data <- read.delim("sampleGroups.txt", header = T, row.names = 1)
head(meta_data)
```
If the row names of metadata and col names of raw counts data:
idx <- match(rownames(metadata), colnames(rawcounts))
and apply the idx to the desired data to be reordered.

## Modelling DeSeq2 Object
```{r}
dds <- DESeqDataSetFromMatrix(countData = raw_counts,
                              colData = meta_data,
                              design = ~group) #group is the condition:resistant or sensitive in this case
dds
```

## Filter out low gene counts
```{r}
keep <- rowSums(counts(dds) >= 10) >= 3
dds <- dds[keep,]

```

counts(dds) -> gets the count matrix
rowSums() -> for each gene counts the samples that are over 10

## DE Analysis
```{r}
dds <- DESeq(dds)
```


## Normalization and Log Transformation (transformed counts)
```{r}
#estimateSizeFactors(dds)
#sizeFactors(dds)

#normalization
norm.counts <- counts(dds, normalized = TRUE) #normalize the counts for accuracy
write.csv(norm.counts, file = "normalized_counts.txt", row.names = TRUE)
head(norm.counts)

#log transformation, transformed counts
vsd <- vst(dds, blind = FALSE) 

#creating as a matrix
vsd_matrix <- assay(vsd)
head(vsd_matrix)
```
## Adding correlation to log transformation
```{r}
vsd_cor <- cor(vsd_matrix) #applies correlation for a pheatmap
head(vsd_cor)
vsd_cor <- data.frame(vsd_cor)

pheatmap(vsd_cor, annotation = select(meta_data, group))
```
1 indicates strong correlation - shows a good biological relevance.
- Same groups have correlation of 1.

This plot supports that:
- Chemo-sensitive and chemo-resistant samples have distinct gene expression profiles -> distinct transcriptomic level.
- The clustering and correlation structure suggest clear biological separation, which is a good foundation for further differential expression or biomarker analysis.

## Unsupervised Clustering
```{r}
plotPCA(vsd, intgroup = c("group")) # does not want the matrix version
```
- PC1 explains 87% of the variance — this is the main axis separating your samples.
- Clear separation between two group -> This suggests strong transcriptomic differences between the two groups.
- PC2 (8% variance) captures minor variation within groups.

## Hierarchical Clustering
```{r}
#plotting hierarchical clustering using vst transformed counts
sampleDists <- dist(t(assay(vsd)))
plot(hclust(sampleDists, method = "complete"))
```

## Plot dispersion estimates -- used DESeq2 object
```{r}
mean.counts <- apply(raw_counts[,1:3],1,mean)

var.counts <- apply(raw_counts[,1:3],1,var)

# Plot dispersion estimates to visualize variability in gene expression
# - Black dots: raw gene-wise dispersion estimates (highly variable for low-count genes)
# - Red line: fitted trend showing expected dispersion based on mean counts
# - Blue dots: final dispersion values used in the model (shrunken for stability)
# This plot helps assess whether DESeq2's dispersion modeling and shrinkage worked correctly

plotDispEsts(dds) 

```

# Key observation:
 - Low-count genes (left) have high raw dispersion, but shrinkage pulls them toward the red trend line.
 - High-count genes (right) show less variability and need minimal shrinkage.
This confirms DESeq2's dispersion estimation and shrinkage worked correctly, improving statistical robustness.
• As mean counts increase, raw dispersions are more stable; blue ≈ black (less shrinkage needed).
• Overall fit looks good (no strong systematic deviations), so dispersion modeling succeeded and you can confidently proceed to differential expression testing.

## Get DE results
```{r}
dds.res <- results(dds, contrast = c("group","chemo_sensitive", "chemo_resistant"), alpha = 0.05)
summary(dds.res)
mcols(dds.res)$description

write.csv(as.data.frame(dds.res), file="DE_sensitiveVSresistant_results.csv")
```
Summary of Differential Expression Results:
- 18,577 genes had a nonzero total read count (i.e., were expressed in at least one sample).
- 2,156 genes were upregulated in chemo_sensitive vs chemo_resistant (12%).
- 1,875 genes were downregulated (10%).
- This means ~22% of all expressed genes are significantly different between the two conditions.
- 8 genes (0.043%) were flagged as outliers due to extreme values (checked using Cook’s distance).
- 0 genes were filtered out due to low mean counts (<6), likely because of independent filtering.

```{r}
plotMA(dds.res)
```
- LFC > 0 is up-regulated genes; LFC < 0 is down-regulated genes.
- Grey points are the genes that are not statistically significant.
- Blue points genes with p.adj < 0.05 (significantly differentially expressed)
- There is a large number of significantly differentially expressed genes, especially in genes with moderate-to-high expression (middle x-axis).
- The triangle markers suggest some genes have very large fold changes and may be biologically interesting.

## Volcano Plot
```{r}
BiocManager::install('EnhancedVolcano')
library(EnhancedVolcano)
```

```{r}
EnhancedVolcano(dds.res,
    lab = rownames(dds.res),
    x = 'log2FoldChange',
    y = 'padj')
```
- genes to the right are up-regulated due to LFC > 0
- genes to the left are down-regulated due to LFC < 0
- higher values are more statisticaly significant
- There are many red points, meaning a large number of genes show both statistically significant changes and large expression differences between the two conditions.
- The volcano plot confirms the biological distinction between chemo-sensitive and resistant samples.
- Genes near the center LFC~0 are not statistically significant.

## Most significant DE genes
```{r}
plotCounts(dds, gene=which.min(dds.res$padj), intgroup="group")
# the lowest padj indicates the most significant gene
```
## Visualizing expression data with a heatmap
```{r}
# extracting significantly DE genes -- padj < 0.05 & abs(lfc) > 1
# thresholding for significanlty differentially expressed genes with at least two-ford change in expression
de.sign <- subset(dds.res, padj < 0.05 & abs(log2FoldChange) > 1)

# creating a vector of Ensembl id for those genes
de.sign.genes <- rownames(de.sign)

# subseting the vsd matrix to include only the de.sign.genes and scaling it by row so that we plot on the heatmap
scale.dat <- t(scale(t(assay(vsd)[de.sign.genes,])))
pheatmap(scale.dat[de.sign.genes,], cluster_rows=TRUE, show_rownames=FALSE, cluster_cols=TRUE)
```

## Annotating DE Results

```{r}
BiocManager::install('AnnotationDbi')
BiocManager::install("org.Hs.eg.db")
library(AnnotationDbi)
library(org.Hs.eg.db)
```
```{r}
res.df <- as.data.frame(dds.res) # putting the de.sign into a data.frame to help the next step to work
res.annot <- mapIds(org.Hs.eg.db, keys = rownames(res.df), keytype = "ENSEMBL", column = "SYMBOL") #mapping ensembl ids to Gene Symbols

res.df.annot <- cbind(res.df,gene.symbol = res.annot) #binding the annotations to the sign.df

head(res.df.annot)

write.csv(res.df.annot, file = "DE_sensitiveVSresistant_results_wGeneSymbol.csv")
```
```{r}
EnhancedVolcano(res.df.annot,
    lab = res.df.annot$gene.symbol,
    x = 'log2FoldChange',
    y = 'padj')
```

# Over-representation test
## Install R packages and load libraries
```{r}
BiocManager::install("clusterProfiler")
library(clusterProfiler)

install.packages("msigdbr")
library(msigdbr)
```
msigdbr is an R package that gives you easy access to gene sets from the Molecular Signatures Database (MSigDB) — a widely used resource for gene set enrichment analysis (GSEA).

## Load the C5 collection from MSigDB i.e. Gene Ontologies
```{r}
go_gene_sets <- msigdbr(species = "human", category = "C5")  #Load C5

msigdbr_t2g <- go_gene_sets %>% dplyr::distinct(gs_name,gene_symbol) %>% as.data.frame() # make a dataframe of the genesets and the corresponding Ensemble gene ids
#t processes a gene set data frame (likely created using msigdbr()), and formats it for enrichment analysis, especially for use with clusterProfiler::enricher().
```
## Load the DE results file and subset for significantly upregulated genes
```{r}
DE.res <- read.csv("DE_sensitiveVSresistant_results_wGeneSymbol.csv", header = TRUE, row.names = 1)
DE.res.sign <- subset(DE.res, padj < 0.05 & log2FoldChange > 1)
```

## Perform enrichment analysis and write results in a .csv file
```{r}
enrichRes <- enricher(gene = DE.res.sign$gene.symbol, TERM2GENE = msigdbr_t2g, pvalueCutoff = 0.05, pAdjustMethod = "BH")
write.csv(enrichRes@result, file = "GO_sensitiveVSresistant_up.csv", row.names = FALSE)
head(enrichRes)
```


# Pre-ranked GSEA
## Installing packages
```{r}
#BiocManager::install("fgsea")
library(fgsea)

#install.packages("data.table")
library(data.table)

BiocManager::install("qusage")
library(qusage)

#install.packages("ggplot2")
library(ggplot2)
```

## Load the Gene Set collection (Note you need to download the gmt file of the collection of your interest from MSigDB)
```{r}
gmt.file <- read.gmt("c5.go.bp.v7.4.symbols.gmt")
```

## Ranking our DE results
```{r}
DE.res.ranked <- DE.res[order(DE.res$log2FoldChange, decreasing = T), ]

DE.ranks <- setNames(DE.res.ranked$log2FoldChange, DE.res.ranked$gene.symbol) # creating a named vector of the ranks
```


## Running fgsea
```{r}

fgseaRes <- fgsea(gmt.file, DE.ranks, minSize=15, maxSize=500)

# writing the results into a file
fwrite(fgseaRes, file = "PrerankedGSEA_gobp_sensitiveVSresistant.txt", sep = "\t", sep2=c("", " ", ""))
```

## Producing an enrichment plot
```{r}
topPathways <- fgseaRes[order(padj)][padj <= 0.05]$pathway
EnPlot <- plotEnrichment(gmt.file[["GOBP_EPITHELIUM_DEVELOPMENT"]],DE.ranks) + labs(title="GOBP_EPITHELIUM_DEVELOPMENT") + theme(text = element_text(size=20))
EnPlot2 <- plotEnrichment(gmt.file[["GOBP_OXIDATIVE_PHOSPHORYLATION"]],DE.ranks) + labs(title="GOBP_OXIDATIVE_PHOSPHORYLATION") + theme(text = element_text(size=20))

EnPlot
EnPlot2

```
X-axis ("rank"):
Genes ranked from most upregulated (left) to most downregulated (right) based on a metric (e.g., log2FC, signal-to-noise).

Black ticks:
Locations of genes from the epithelium development gene set along the ranked gene list.

Green curve:
The enrichment score (ES) — it reflects how concentrated the gene set is near the top (left) or bottom (right) of the ranked list.

If the peak is on the left: gene set is enriched in upregulated genes.

If on the right: enriched in downregulated genes.

Red dashed lines:
Indicate the range of observed ES scores; useful for permutation testing.

✅ Interpretation of your plot:
The green line peaks early, around the left side of the ranked gene list.

This indicates that genes in the GOBP_EPITHELIUM_DEVELOPMENT gene set are enriched at the top of the ranking.

Therefore, this gene set is upregulated (i.e., activated or associated) in your chemo-sensitive group (if that's the direction of the ranking).

🧠 Biological interpretation:
Epithelium development–related genes are significantly enriched among the most upregulated genes in your condition (likely chemo-sensitive).

This may suggest that epithelial differentiation or maintenance is more active in chemo-sensitive samples.

